Maximum Entropy Spectral Analysis: a case study

### 2023-07-25 paper/mesa.tex
Removed garbage from tex & made new structure

% comments command
\newcommand{\wdp}[1]{{\textcolor{magenta}{\texttt{WDP: #1}} }}
\newcommand{\amartini}[1]{{\textcolor{blue}{\texttt{AM: #1}} }}
\newcommand{\sschmidt}[1]{{\textcolor{red}{\texttt{SS: #1}} }}

@@

\section{Introduction}

The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

After a short introduction to Jaynes' MAXENT (sec.~\ref{sec:MAXENT}), we will develop in detail Burg's technique of Maximum Entropy Spectral Analysis (MESA) and show that the estimate can be expressed in an analytical closed form (sec.~\ref{sec:MESA}).
Next, we will discuss an interesting link between Burg's method and autoregressive processes (sec.~\ref{sec:autoregr}) and in sec.~\ref{sec:forecasting} we will use such link for straightforwardly forecasting a time series.

@@

@@

The remainder of this section is devoted to an extensive study on how to make such choice.
In Section~\ref{sec:optimizers}, we are going to define three different \textit{loss functions} to measure how well the 
algorithm is able to reproduce a known PSD.
The basic idea is to validate, as the autoregressive order considered increases, the performance of the algorithm results 
by measuring the loss function and pick, among the orders the one that yields better results.
Second, the performance of different loss functions will be assessed by performing the spectral analysis on time series with an analytical Gaussian PSD, sec.~\ref{sec:arp_validation}.
In a last subsection~\ref{sec:LIGO_validation}, the same analysis is performed on synthetic data generated from the analytical spectrum released by the LIGO-Virgo collaboration.

@@

Guided from numerical experiments, an indication on the upper bound to the autoregressive order $M_{max}$ is~\citep{doi:10.1190/1.1440902}:
\begin{equation}\label{eq:MMAx}
M_{max} = 2N / \ln{(2N)}\,,
\end{equation}
where $N$ is the number of observed points in the time-series.
However, this is just a plausible upper limit on the order of the AR process $m$ and the optimal algorithm could employ fewer points.
We then need a more sophisticated method for computing the right value for $m$.
Various loss functions to assess the algorithm performance have been proposed in literature.
We summarise them below:

@@

\item \textbf{Criterion Autoregressive Transfer function (CAT)}
This second loss function has been proposed by Parzen and studied in detail by~\citet{bhansali1986}. 
It is based on the assumption that the observed process is an infinite order autoregressive process 
\begin{equation} 
x_t = \sum_{i = 1}^{\infty}a_{i}x_{t-i} + \nu_t
\end{equation} 
and tries to select the order $m$ as the best finite-order approximation for the observed process. Being N the number of samples it has the property that $m \to \infty$ as $N \to \infty$. Since any real-valued stochastic process can be written as a ARMA(p,q) process (Wold theorem) i.e. an $AR(\infty)$ process, this is a physically significant property. 
The loss function has the functional form: 
\begin{equation}
    \mathcal{L}_{\rm CAT}(m) = \frac{1}{N}\sum_{k = 1}^m \frac{N - k}{N P_k} - \frac{N - m}{N P_m},
\end{equation}
and the so-chosen order is found to be asymptotically unbiased for $P_N$.

\item \textbf{Optimum Bayes Decision rule} 
The last criterion we will consider is the Optimum Bayes Decision rule (OBD)\citep{doi:10.1029/WR018i004p01097}. Let $x(t)$ be the observed 
time series for the process that has to be described as an $AR(m)$ process, with $m$ to be determined. The OBD is obtained choosing between M different hypothesis $H_i, i = 1, \dots, M$ maximizing their posterior distribution $P(H_i\vert x(t))$. Each
hypothesis is uniquely determined from both the length and the values of the filter under study. Choosing a Likelihood of the form Eq.~\eqref{eq:p_forecast} and uniform priors for both the coefficients and the hypotheses, Rao has shown that choosing the minimum for $-log(P(H_i \vert x(t))$, i.e. maximizing the posterior for $H_i$, is asymptotically equivalent to the minimization of:

   \begin{align}
        \mathcal{L}_{\rm OBD}&(m) = (N - m - 2) \log(P_m) \nonumber\\
        &+ m \log(N) + \sum_{k = 0}^{m-1} \log(P_k) + \sum_{k = 1}^{m} a_k^2\,.
    \end{align}
\end{itemize}

@@

\subsection{Choice of the loss function: Gaussian PSD} \label{sec:arp_validation}
We test the performance of the three loss functions on a random time series generated with a known power spectral density.
In this first experiment, we take the PSD to be a Gaussian distribution with mean $\mu = 2.5$ and standard deviation $\sigma = 0.5$, where the units are arbitrary.
The time series are generated by sampling a frequency vector from ${p(\tilde{n}(f_i)) \propto \exp{\left\{-\frac{1}{2}\sum_i\frac{n_i ^2}{S(f_i)}\right\}}}$ and performing an inverse Fast Fourier Transform on the frequency series. 
For this investigation, we generate a dataset of $N=1000$ time series of $3000$ points each.

We then apply Burg's method choosing in turn each of the loss functions on the ensemble of simulated time series, 
thus obtaining an ensemble of PSD estimates. Through these ensembles, we characterize statistically the performance of each loss function.
The disagreement between the PSD $S_i(f)$ estimated from the $i$-th simulated time series and the target PSD $S(f)$ is measured via 
the \textit{frequency-averaged relative error} $r_i$:
\begin{equation}\label{eq:freq_error}
	r_i = \frac{1}{N_f}\sum_{f_j=\{0, \hdots , Ny\}} \frac{|S_i(f_j) - S(f_j)|}{S(f_j)}
\end{equation}
where $Ny$ is the Nyquist frequency of the time series and $N_f$ is the number of the discrete frequencies the PSD is evaluated at.

For each loss function, we compute the ensemble-averaged PSD (as well as the $90\%$ confidence level) and 
the \textit{ensemble-averaged relative error}:
\begin{equation}\label{eq:ensemble_averaged_error}
	r(f) = \frac{1}{N} \sum_i \frac{|S_i(f) - S(f)|}{S(f)} \;.
\end{equation}
For each loss function, Figs.~\ref{fig:FPEmean} \ref{fig:OBDmean} and \ref{fig:CATmean} show the averaged PSD, 
as well as the ensemble-averaged relative error from Eq.~\eqref{eq:ensemble_averaged_error}.
Furthermore, Fig.~\ref{fig:optcomparison} displays the relation between $r_i$ and the autoregressive 
order $m_i$ chosen for each independently drawn time series.

\begin{figure}[t]
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/FPE_spectrum_estim.pdf}
	\caption{In the top panel, we show the average spectrum for $\mathcal{L}_{\rm FPE}(m)$ with 90\% confidence regions (purple shaded region), together with the median estimation (red line) and the target PSD (black dashed line). In the bottom panel, we display the ensemble-averaged relative error eq.~\eqref{eq:ensemble_averaged_error}.
	The average is computed with $1000$ realization of a $3000$ points long time series.}
	\label{fig:FPEmean}
\end{figure}
\begin{figure}[t]
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/OBD_spectrum_estim.pdf}
	\caption{In the top panel, we show the average spectrum for $\mathcal{L}_{\rm OBD}(m)$ with 90\% confidence regions (purple shaded region), together with the median estimation (red line) and the target PSD (black dashed line). In the bottom panel, we display the ensemble-averaged relative error eq.~\eqref{eq:ensemble_averaged_error}. The average is computed with $1000$ realization of a $3000$ points long time series.}
	\label{fig:OBDmean}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/CAT_spectrum_estim.pdf}
	\caption{In the top panel, we show the average spectrum for $\mathcal{L}_{\rm CAT}(m)$ with 90\% confidence regions (purple shaded region), together with the median estimation (red line) and the target PSD (black dashed line). In the bottom panel, we display the ensemble-averaged relative error eq.~\eqref{eq:ensemble_averaged_error}. The average is computed with $1000$ realization of a $3000$ points long time series.}
	\label{fig:CATmean}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width = \linewidth]{Images/optimisers_comparison/normal/error_length_contour.pdf}
	\caption{Frequency-averaged relative error $r$ eq.~\eqref{eq:freq_error} vs the length $m$ of autoregressive process, for each of the $1000$ independent realizations of the synthetic time series with a gaussian PSD. Different colours refers to different choices for the loss function: in red $\mathcal{L}_{\rm FPE}(m)$, in green $\mathcal{L}_{\rm OBD}(m)$ and in black $\mathcal{L}_{\rm CAT}(m)$. The top and right histograms show the marginal distributions.}
	\label{fig:optcomparison}
\end{figure}

\paragraph{Final Prediction Error (FPE)}
The results of our investigation on the performance of FPE are shown in Fig.~\ref{fig:FPEmean}.
Qualitatively, there is a good agreement between the reconstructed spectrum and the true spectrum; 
however, we note that the reconstruction is not very accurate in the low frequency region. 
Furthermore, the $90\%$ credible region is very small: this means that if we randomly take two of the reconstructed spectra, we expect their differences to be statistically small. These facts are evidence for FPE to be a reliable loss function.

By looking at Fig.~\ref{fig:optcomparison} (red series), we note that the AR orders obtained with 
FPE are clustered in a very small region. This is also a desirable property: FPE, in fact, provides a stable 
estimation of the AR order, which does not affect much the reconstruction error.
We conclude that the FPE shows good quality reconstruction for the spectrum and very desirable stability properties, its estimate for filter's length $m$ is clustered in a region where there is no dependence of error on $m$. 
\paragraph{Optimum Bayes Decision Rule (OBD)}
The second loss function we consider is OBD and the results are summarized in Fig.~\ref{fig:OBDmean}. As for the FPE case, they show a good agreement between the average over the reconstructions and the true spectrum.
Qualitatively the same behaviour of FPE is observed: a good quality reconstruction at the intermediate and high frequencies with a narrow $90\%$ confidence level as well as a degrading performance at low frequencies.
However, when looking at the error, the disagreement of this method is found to be larger compared to FPE: 
in the worst case, the error can be as large as a factor of $2$ compared to FPE.

The error as a function of the AR length (green series in Fig.~\ref{fig:optcomparison}) clusters in a small region, indicating the stability of the reconstructed process order $m$. We note that on average, OBD tends to choose smaller values of $m$ with respect to FPE.

\paragraph{Criterion Autoregressive Transfer function (CAT)}
The performance of CAT is shown in Fig.~\ref{fig:CATmean}. At a first glance, CAT differs substantially from the other two loss functions considered.
The ensemble-average PSD matches very well the underlying ``true" PSD. This is also true even in the low frequency region, 
where both OBD and FPE showed poor performance.
However, the variance of the reconstructed spectrum is quite large (much larger than for FPE and OBD), and the relative error is quite high, $\sim 10\%$ and it is approximately constant over all the frequencies.

The reason for this behaviour becomes clear from Fig.~\ref{fig:optcomparison} (black series): CAT does not converge 
to any specific value for the order of the AR process. The estimated $m$ spans a large range of values, hence the large variance observed.
Fig.~\ref{fig:optcomparison} also shows a strong dependence of the error on the estimated length of the filter. The good quality of the reconstruction from the average spectrum can be explained as follows: long filters are able to capture features that short filters cannot see, like outliers in different realizations of the time-series, but this is also responsible for an increased variance in the estimate, by introducing spurious peaks in the reproduction.

When averaging the different PSD estimates, the noise in each spectrum cancels, as expected from the Gaussian nature of the AR process. 
This implies, in a sense, that each estimate of the spectrum is independent of any other, as suggested by the huge variance in the residuals. This lack of stability is not a good property for the estimation of the PSD from a single realization of the time-series, however, thanks to the averaging out of errors, this estimator seems optimal in the case of repeatable experiments and ensemble-averages.

\paragraph{Final remarks on the choice of the loss function}

In our analysis, the FPE and OBD loss functions are found to behave similarly while CAT shows fairly different properties.
CAT provides an accurate average spectrum over all the frequencies at the price of a large variance; in turn OBD and FPE 
provide a poorer average in the low frequency tails, however they also display a  smaller variance, with FPE showing the lowest.

The poor low-frequency reconstruction from OBD and FPE might be due to the fact that the first tends to select shorter filters, whereas long filters are required to model low frequency correlation.
This seems to be confirmed by looking at fig.~\ref{fig:optcomparison}. OBD, which select the shortest filters, can provide an error as large as $300 \%$ at the extrema, as compared with $85\%$ error of FPE. In turn, CAT is $5\%$ accurate in these regions.

However, we also note that the when inferring PSDs from a single time-series realization, FPE provide the lowest averaged error over all frequencies, while CAT can reach errors $5$ times larger (see again Fig.~\ref{fig:optcomparison}).
Hence, while CAT is the loss function that minimizes errors in the low-frequency end of the spectrum, FPE obtains the best overall accuracy.

The conclusion is clear: even if in some cases CAT is more accurate when taking the average over several realizations of the underlying process, 
FPE guarantees that the single estimation is more faithful.
As in any common situation we cannot perform such averaging over different realizations of the same time series, we must prefer FPE over CAT (let alone OBD, which even though qualitatively similar to FPE has worse performance).
However, if we indeed can measure the PSD by averaging over different time series, using CAT as a loss function is the best choice. In this sense we retain CAT to provide the best, and most similar in spirit, alternative to the commonly employed Welch estimation method whenever ensemble-averages are needed and justified.

\subsection{Choice of the loss function: LIGO Spectrum} \label{sec:LIGO_validation}
We continue our characterization of the various loss functions considered in this work, by investigating the 
reconstruction of a specific, known, power spectrum: that is the Advanced LIGO \citep{LIGO_2015, Virgo_2014, KAGRA_2012, KAGRA_2013} design sensitivity theoretical spectral curve \citep{LVK_future_2013, LIGO_PSDs}.
For this analysis, we generate $N = 500$ time series of $40960$ points each, sampled with a sampling rate of $\SI{2048}{Hz}$, hence we fix the duration
of the time-series to $\SI{20}{s}$. The chosen length is convenient to capture fairly accurately the low-frequency features of the LIGO PSD.
We report our findings in Figs.~\ref{fig:ligospectrum}~and~\ref{fig:LigoOrderError}.

\begin{figure}
    \centering
     \includegraphics[width = \linewidth]{Images/optimisers_comparison/ligo/compare_estimates.pdf}
      \caption{Average spectrum for the three different loss functions as compared with the ``true" PSD. The average is computed over $500$ realization of a $40960$ points long time series.}
       \label{fig:ligospectrum}
\end{figure}

Fig.~\ref{fig:ligospectrum} shows the simulated spectrum (dashed line) and the ensemble-averaged reconstructed PSD adopting the FPE (green line), OBD (red line) and CAT (black line).
In all cases, the spectrum is well reconstructed, but with a fairly distinct behaviour at low frequency, where CAT -- as in the Gaussian case -- better captures and resolves the distinct spectral feature at $\sim \SI{17}{Hz}$.
In Fig.~\ref{fig:ligoPeaks}, we report the reconstructed spectra around the two peaks at $\sim \SI{17}{Hz}$ and $\sim \SI{438}{Hz}$ (left and right panel respectively).

Fig.~\ref{fig:LigoOrderError} shows the distribution of recovered AR orders $m$ against the relative frequency-averaged error. The behaviour
of the three loss functions is very similar to what found in the Gaussian PSD case (compare it with fig.~\ref{fig:optcomparison}): ODB infers the smallest orders and gives average errors around $20$\%, FPE consistently estimates orders of a few hundreds and shows the smallest errors $\sim 15$ \% while CAT does not show any preference towards any AR order and displays wildly varying errors. Yet again, when the PSD is averaged over multiple realization of the time, CAT is able to capture the spectrum very precisely. In fact, even in presence of very sharp spectral features, CAT reconstruction seems to be almost perfectly coincident with each of them.
Hence, also the study of simulated LIGO data seems to indicate that whenever and wherever ensemble-averaged PSDs are necessary, CAT is the optimal choice of loss function. However, on a single time-series realization, FPE is the more robust choice.

Let us summarize some key general conclusions:
\begin{itemize}
	\item there are no reasons to prefer OBD over CAT or FPE;
	\item if we have one single realization for the process, we recommend the use of FPE, that would get the best resolution possible. In this situation, CAT would provide spurious and unreliable results, with large error;
	\item in the case of several realizations of the same process, CAT ensemble-average properties provide very a precise spectral estimation.\end{itemize}

Therefore, the choice of loss function, at least in between CAT and FPE, depends on the problem one is attempting to solve.
\begin{figure}
    \centering
    \includegraphics[width = \linewidth]{Images/optimisers_comparison/ligo/error_length_contour.pdf}
    \caption{For each of the $500$ independent realization of the time series, we plot the relative error $r$ (as in eq.~\eqref{eq:freq_error}) against the length $m$ of autoregressive process. The time series are randomly drawn with a the analytical LIGO PSD in fig.`\ref{fig:ligospectrum}. Different colors refers to different choices for the loss function. Histograms for the distribution od the individual quantities are also represented.}
    \label{fig:LigoOrderError}
\end{figure}

\begin{figure}[t]
        \includegraphics[width = \linewidth]{Images/optimisers_comparison/ligo/compare_estimates_peaks.pdf}
        \caption{Details of peaks of the spectrum and their reconstruction with every optimizer}
        \label{fig:ligoPeaks}
\end{figure}

\subsection{How well is the AR order recovered?}

@@

We now address the issue of how well the AR order (i.e. the number of $a_k$ coefficients employed) is estimated by each loss function.
This is useful to further characterize the properties of the different loss functions. Furthermore, as MESA is supposed to model data as an $AR(p)$ process, it is interesting to quantify its precision on some true autoregressive (and stationary) time series.

We generate $100$ autoregressive processes $AR(p)$ with a random value of $p$, drawn such that $log(p) \sim\mathcal{U}_{[log(2), log(5000)]}$.
Each coefficient $a_k$ is assigned according to a Dirichlet distribution $\mathrm{Dir}([1,\hdots, 1])$. The sign of $a_k$ is assigned randomly according to a binomial distribution. We report the result of this investigation in Fig.~\ref{fig:p_vs_ptrue}.

It can be seen that OBD and FPE loss functions show very similar behaviour. They are {\it very} reliable in capturing the correct AR order up to a certain threshold ($p\sim 100$ for OBD and $p\sim 200$ for FPE). For any AR process of order higher than the threshold, the optimization underestimates badly the actual value of $p$.
For this reason, FPE and OBD seem reliable only for relatively simple AR processes. For more complicated processes, they seem to ``underfit'' the problem (i.e. they output a model simpler that those required).

On the other hand, CAT shows a different behaviour. The selected AR order $p$ is always close to the maximum possible value, regardless the actual value $p_{true}$ of the underlying process. This produces always a model that is more complex than those obtained with FPE and OBD and this can explain the origin of the high variance in the estimation observed in the experiments described in the previous sections.
Of course, CAT does not perform a good job in reconstructing the AR process. However, for high order AR processes, the error introduced by CAT is more tolerable than the error introduced by FPE and OBD. This is a good reason to prefer CAT in these situations (see also Sec.~\ref{sec:temp_timeseries} for another example of this effect).

\section{Comparison with Welch method}\label{sec:Welch_comparison}
We perform a {\it qualitative} comparison between the performance of the MESA and of the standard Welch algorithm.
In this, we cannot avoid to be only qualitative. Indeed, as the results of the comparison are problem dependent, it is very hard to quantify this in a single metric. Although similar studies can be drawn from any other PSD, in this section we focus on a single PSD and we try to generalize some observations that we make.
We decide to use the analytical PSD computed for LIGO Handford interferometer, released together with the GWCT-1 catalog~\citep{GWTC1, PSD_release}, and computed with BayesLine package \citep{Cornish_2015, Littenberg_2015, Cornish_2020, Chatziioannou_2019}.

We simulate data\footnote{This is to ensure that we have a baseline PSD to compare the data with} from the PSD used for the analysis of the event GW150914 and we employ both Welch's method and MESA to estimate the spectrum.
We vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we choose the FPE loss function. For the Welch algorithm, we employ a Tukey window with a parameter $\alpha$ 
equal to 0.4, an overlap fraction of $1/2$ for the segments and a length of 
segments $L = 512, 1024, 2048, 8192, 32768$ points, depending on the observation time.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.
For the Welch algorithm, we use the standard implementation provided by the python library \texttt{scipy}~\citep{numpy,scipy}.
The results from both methods are summarized in  Figs.~\ref{fig:MESA_LIGO_data} and~\ref{fig:welch_LIGO_data} respectively.

@@

\section{Applications} \label{sec:applications}
%\subsection{Temperature Time Series} \label{sec:temp_timeseries}
%As a further example of the breadth of applicability of MESA, we applied our implementation to atmospheric temperature time series. 
%The reason for this choice is twofold: i) atmospheric temperature time series present a variety of overlapping periodicities most of which are 
%known; ii) it provides a stress test for the time series forecast analysis.  As dataset, we  used the historical reanalysis data 
%from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store~\citep{ERA5_dataset}.
%The data consist in temperatures taken at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperatures are given on an hourly cadence for almost 31 years from 31st December 1989 to 30th November 2020.
%
%Fig.~\ref{fig:temp_spectrum}, shows the MESA spectrum inferred from the data. As a comparison, we adopt both the FPE and the CAT loss functions.
%In agreement with what we found in previous sections, the FPE PSD is more regular compared to the one from CAT. Both spectra show a
%peak at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to one day, corresponding to the day-night cycle: this is expected.
%Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they correspond to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.
%
%Looking at low frequencies, FPE does not capture the yearly variability at $f_{yr} = \SI{1}{yr^{-1}}$. On the other hand, the peak is captured well by the CAT loss function. In analogy as what observed for the peak at $f_D$, in the CAT spectrum we observe a higher order harmonic at a frequency $2 \cdot f_{yr}$. As in the previous case, this is required to better model the temperature variation in the year.
%
%The failure of FPE in capturing the low frequency trend can be understood by looking at Fig.~\ref{fig:p_vs_ptrue}. It is shown that CAT systematically select very long filters (i.e. large $p_{CAT}$ values for the autoregressive orders), whereas FPE tends to select smaller values for the order $p_{FPE}$ of the autoregressive process, often underestimating the actual value.
%In this example, we have $p_{CAT} \sim  36500 \sim  \SI{4}{yr}$, whereas $p_{FPE} \sim  1300 \sim  \SI{2}{months}$.
%The autoregressive process selected by FPE, hence cannot produce a meaningful prediction on the timescale longer than a few weeks and for this reason it is unable to capture the low frequency peak at $f_{yr}$. In other words, the model is too simple to model \textit{both} the behaviour at high and low frequencies (separated by approximately 3 orders of magnitude).
%On the other hand, the model chosen by CAT is much more complex and is able to model also the high frequency behaviour, at the expense of making a more noisy estimation.
%The ``actual" length of the autoregressive process should be closer to the choice of CAT.
%
%We now assess the accuracy of the forecasting of new observations of the time series. 
%Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ independent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temperature values. The predictions span a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.
%
%We note the observed difference is always well included in the $90\%$ confidence interval: the forecasting predictions seem reliable.
%On the other hand, the confidence interval is pretty large (almost as large as $\SI{15}{K}$), thus making ``easy" for the actual data to fit the predictions.
%Indeed, the prediction model, while suitable for spectral estimation, is nothing more than a linear regression (plus noise term). Such a simple model hardly catches the complex trend in the variability of the temperature daily trend during a year. For more precise predictions, probably one should consider nonlinear regression model, tapping into the wealth of non linear predictors offered by the field of Deep Learning.
%
%\begin{figure}
%	\caption{Spectrum for the historical temperature time series. The two lines refers to different loss functions (CAT and FPE); the inset shows the harmonics of the fundamental frequency of a day. Two vertical lines are drawn in correspondence to the frequency $f_D$ of a day and $f_{yr}$ of a year.}
%	\label{fig:temp_spectrum}
%	\includegraphics{Images/climate_plots/temp_spectrum.pdf}
%\end{figure}
%
%\begin{figure}
%	\caption{Difference between the historical temperature $T$ and the forecast temperature $T_{forecast}$ for two years of data. The red shadow denotes the $90\%$ confidence interval of the predictions. The model is trained with CAT loss function. The discrepancy between predicted and actual values is always everywhere included in the $90\%$ confidence interval.}
%	\label{fig:temp_forecast}
%	\includegraphics{Images/climate_plots/forecast_accuracy.pdf}
%\end{figure}

>>>>

% comments command
\newcommand{\wdp}[1]{{\textcolor{magenta}{\texttt{walter: #1}} }}
\newcommand{\amartini}[1]{{\textcolor{blue}{\texttt{alessandro: #1}} }}
\newcommand{\sschmidt}[1]{{\textcolor{purple}{\texttt{stefano: #1}} }}
\newcommand{\todo}[1]{{\textcolor{red}{\texttt{TODO: #1}} }}

@@

\section{Introduction}
\todo{Check again intro \& abstract}
The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

@@

After a short introduction to Jaynes' MAXENT (Sec.~\ref{sec:MAXENT}), we will develop in detail Burg's technique of Maximum Entropy Spectral Analysis (MESA) and show that the estimate can be expressed in an analytical closed form (Sec.~\ref{sec:MESA}).
Next, we will discuss an interesting link between Burg's method and autoregressive processes (Sec.~\ref{sec:autoregr}) and in Sec.~\ref{sec:forecasting} we will use such link for straightforwardly forecasting a time series.

@@

\subsection{Whitening} \label{sec:whitening}
\todo{Write here about our whitening procedure with the convolution}

@@

The remainder of this section is devoted to an extensive study on how to make such choice.
In Section~\ref{sec:optimizers}, we are going to define two different \textit{loss functions} to measure how well the 
algorithm is able to reproduce a known PSD.
The basic idea is to validate, as the autoregressive order considered increases, the performance of the algorithm results 
by measuring the loss function and pick, among the orders the one that yields better results.
The performance of the different losses will be assessed by answering to two questions: (i) how well the AR order is recovered and (ii) how well the measured PSD is able to whiten the input time series.
This will be discussed Sec.~\ref{sec:arp_validation} and Sec.~\ref{sec:whitening_validation}.

@@

Guided from numerical experiments, an indication on the upper bound to the autoregressive order $M_{max}$ is~\citep{doi:10.1190/1.1440902}:
\begin{equation}\label{eq:MMAx}
M_{max} = 2N / \ln{(2N)}\,,
\end{equation}
where $N$ is the number of observed points in the time-series.
However, this is just a plausible upper limit on the order of the AR process $m$ and the optimal algorithm could employ fewer points.
We then need a more sophisticated method for computing the right value for $m$.
We summarise them below:

@@

\item \textbf{Variance Maximum (VM)}
\todo{WRITEME}
\end{itemize}

@@

\subsection{How well is the AR order recovered?}\label{sec:arp_validation}
\todo{WRITEME}

@@

We now address the issue of how well the AR order (i.e. the number of $a_k$ coefficients employed) is estimated by each loss function.

\subsection{How well can MESA whiten the data?}\label{sec:whitening_validation}
\todo{WRITEME}

\section{Comparison with Welch method}\label{sec:Welch_comparison}
\todo{Check again the section}
We perform a {\it qualitative} comparison between the performance of the MESA and of the standard Welch algorithm.
In this, we cannot avoid to be only qualitative. Indeed, as the results of the comparison are problem dependent, it is very hard to quantify this in a single metric. Although similar studies can be drawn from any other PSD, in this section we focus on a single PSD and we try to generalize some observations that we make.
We decide to use the analytical PSD computed for LIGO Handford interferometer, released together with the GWCT-1 catalog~\citep{GWTC1, PSD_release}, and computed with BayesLine package \citep{Cornish_2015, Littenberg_2015, Cornish_2020, Chatziioannou_2019}.

We simulate data\footnote{This is to ensure that we have a baseline PSD to compare the data with} from the PSD used for the analysis of the event GW150914 and we employ both Welch's method and MESA to estimate the spectrum.
We vary the length of the data used for the estimation: this is also useful to assess how the computation depends on the data available. We set the total observation time $T = 1, 5, 10, 100, 1000 \SI{}{s}$
For the MESA algorithm, we choose the ???? loss function. For the Welch algorithm, we employ a Tukey window with a parameter $\alpha$ 
equal to 0.4, an overlap fraction of $1/2$ for the segments and a length of 
segments $L = 512, 1024, 2048, 8192, 32768$ points, depending on the observation time.
In all cases, the sampling rate is set to $\SI{4096}{Hz}$.
For the Welch algorithm, we use the standard implementation provided by the python library \texttt{scipy}~\citep{numpy,scipy}.
The results from both methods are summarized in  Figs.~\ref{fig:MESA_LIGO_data} and~\ref{fig:welch_LIGO_data} respectively.

@@

\section{An application: marginalising over the noise PSD in GW parameter estimation} \label{sec:applications}
%\subsection{Marginalising over the noise PSD in GW parameter estimation: the case for GW150914}
Since the order of the AR process is not yet determined by the theory, up to this point, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in Sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance and better accuracy than Welch algorithm.
The use of MESA is particularly useful for short time series samples, where Welch's method is outperformed in both precision and confidence.

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\todo{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{|\tilde{d}(f)- \tilde{x}(f;\theta)|^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \citep{R_ver_2010, R_ver_2011, Edwards_2020, Chatziioannou_2021}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \citep{Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\citep{martini_thesis}. Other studies have been done in this direction, mostly using a parametric model for the PSD \citep{Littenberg_2013, Edwards_2015, lalinference}.
